---
title: "newSimu"
output: html_document
date: "2023-01-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Based on F
What we have is 
$$\frac{\hat{R}_{Xg}^2 / M}{(1 - \hat{R}_{Xg}^2) / (n_1 - M  -1)}\sim F_{M, n_1 - M - 1}(\lambda_{Xg})$$
$$\frac{\hat{R}_{Yg}^2 / M}{(1 - \hat{R}_{Yg}^2) / (n_2 - M  -1)}\sim F_{M, n_2 - M - 1}(\lambda_{Yg})$$
where 
$$\lambda_{Xg} = \frac{\boldsymbol{\beta}_{Xg}^T\mathbf{G}_1^T\mathbf{G}_1\boldsymbol{\beta}_{Xg}}{\sigma_X^2}$$
and 
$$\lambda_{Yg} = \frac{\beta_{YX}^2\boldsymbol{\beta}_{Xg}^T\mathbf{G}_1^T\mathbf{G}_1\boldsymbol{\beta}_{Xg}}{\sigma_Y^2 + \beta_{YX}^2\sigma_X^2}$$
For large $n_1/n_2$ and $M$, the following is given 
$$\frac{F_{M, n_1 - M - 1}(\lambda_{Xg}) - \frac{(n_1 - M - 1)(M + \lambda_{Xg})}{M(n_1 - M - 3)}}{\frac{n_1 - M - 1}{M}\left[\frac{2}{(n_1 - M - 3)(n_1 - M - 5)}\left(\frac{(M + \lambda_{Xg})^2}{n_1 - M -3} + M + 2\lambda_{Xg}\right)\right]^{1/2}}\sim N(0, 1)$$

```{r}
genData <- function(oriSNP, u.sd, num.sample, sample.type = 'sample', mdl2.type, sigma_Y, 
                    stage1_mdl, ori.mdl2, stage2_mdl, stage2_mdl_null, effect.size.U, effect.size.X, split.r = 1, setseed = 1025) {
  # get original dataset size
  n0 <- nrow(oriSNP)
  
  # sample U ~ N(0, hat_u_sd)
  set.seed(setseed)
  
  # sample with replacement to get SNPs
  if(sample.type == 'sample') {
    nall <- num.sample
    SNPs_idx <- sample(n0, num.sample, replace = TRUE)
    SNPs <- oriSNP[SNPs_idx, ]
  }else{
    SNPs <- NULL
    for (i in 1:num.sample) {
      SNPs <- rbind(SNPs, oriSNP)
    }
    nall <- num.sample * n0
  }
  U <- rnorm(num.all, sd = u.sd)
  
  # generate X
  # include.idx <- sapply(strsplit(rownames(summary(stage1_mdl)$coefficients), "SNP")[-1], function(x) as.numeric(x[2]))
  # X <- as.matrix(cbind(1, SNPs[, include.idx])) %*% matrix(stage1_mdl$coefficients, ncol = 1) + U
  X <- predict(stage1_mdl, newdata = data.frame(SNPs)) + U
  
  # generate y
  if (ori.mdl2 == "GLM") {
    if (effect.size.X == 0) {
      logit.p <- cbind(1, U) %*% matrix(stage2_mdl_null$coefficients * c(1, effect.size.U), ncol = 1) + rnorm(num.sample, 0, sigma_Y)
    }else{
      logit.p <- cbind(1, X, U) %*% matrix(stage2_mdl$coefficients * c(1, effect.size.X, effect.size.U), ncol = 1) + rnorm(num.sample, 0, sigma_Y)
    }
    p <- exp(logit.p) / (1 + exp(logit.p))
    y <- rbinom(num.sample, 1, p)
  }else if (ori.mdl2 == "LM") {
    if (effect.size.X == 0) {
      y <- cbind(1, U) %*% matrix(stage2_mdl_null$coefficients * c(1, effect.size.U), ncol = 1) + rnorm(num.sample, 0, sigma_Y)
    }else{
      y <- cbind(1, X, U) %*% matrix(stage2_mdl$coefficients * c(1, effect.size.X, effect.size.U), ncol = 1) + rnorm(num.sample, 0, sigma_Y)
    }
  }

  # get datasets for two stages
  n1 <- floor(nall / (1 + split.r))
  idx1 <- sample(nall, n1)
  idx2 <- setdiff(1:nall, idx1)

  y1 <- scale(y[idx1], scale = FALSE)
  y2 <- scale(y[idx2], scale = FALSE)
  X1 <- scale(X[idx1], scale = FALSE)
  X2 <- scale(X[idx2], scale = FALSE)
  SNP1 <- scale(SNPs[idx1, ], scale = FALSE)
  SNP2 <- scale(SNPs[idx2, ], scale = FALSE)

  
  stage1.df.XtoY <- data.frame(Y = y1, X = X1, SNP = SNP1)
  stage2.df.XtoY <- data.frame(Y = y2, X = X2, SNP = SNP2)
  stage1.df.YtoX <- data.frame(Y = X1, X = y1, SNP = SNP1)
  stage2.df.YtoX <- data.frame(Y = X2, X = y2, SNP = SNP2)
  
  colnames(stage1.df.XtoY)[3:ncol(stage1.df.XtoY)] <- colnames(oriSNP)
  colnames(stage2.df.XtoY)[3:ncol(stage2.df.XtoY)] <- colnames(oriSNP)
  colnames(stage1.df.YtoX)[3:ncol(stage1.df.YtoX)] <- colnames(oriSNP)
  colnames(stage2.df.YtoX)[3:ncol(stage2.df.YtoX)] <- colnames(oriSNP)

  return(list(XtoY.df = list(stage1.df.XtoY, stage2.df.XtoY), 
              YtoX.df = list(stage1.df.YtoX, stage2.df.YtoX))
}


fTest2 <- function(rx2, ry2, M, n1, n2, lbd1, lbd2) {
  ## get the statistics
  statx <- rx2 / (1 - rx2)
  staty <- ry2 / (1 - ry2)
  
  varx <- 2 / (n1 - M - 3) / (n1 - M - 5) * ((M + lbd1)^2 / (n1 - M - 3) + M + 2 * lbd1)
  vary <- 2 / (n2 - M - 3) / (n2 - M - 5) * ((M + lbd2)^2 / (n2 - M - 3) + M + 2 * lbd2)
  
  diff_stat <- (statx - staty) / sqrt(varx + vary)
  pval_xy <- 1 - pnorm(diff_stat)
  
  return(list(test.stat = diff_stat, pval = pval_xy))
}

glmR2_func <- function(y, phat) {
  SST <- sum((y - mean(y))^2)
  SSE <- sum((y - phat)^2)
  Rss2 <- 1 - SSE / SST
  Rr2 <- sum((y - mean(y)) * (phat - mean(y)))^2 / (sum((y - mean(y))^2) * sum((phat - mean(y))^2))
  return(c(Rss2, Rr2))
}


R2Test <- function(rx2, ry2, M, n1, n2, alpha, type, y, yhat) {
  if(type == "LM") {
    # var1 <- 4 * rx2 * (1 - rx2)^2 * (n1 - M - 1)^2 / ((n1^2 - 1) * (n1 + 3))
    # var2 <- 4 * ry2 * (1 - ry2)^2 * (n2 - M - 1)^2 / ((n2^2 - 1) * (n2 + 3))
    var1 <- 4 / n1 * rx2 * (1 - rx2)^2 * (1 -  (2 * M + 5) / n1)
    var2 <- 4 / n2 * ry2 * (1 - ry2)^2 * (1 -  (2 * M + 5) / n2)
    estsd <- sqrt(var1 + var2)
  }else{
    pp <- mean(yhat)
    V1 <- mean(yhat * (1 - yhat))
    V2 <- var(y)
    # V2 <- pp * (1 - pp)
    Z <- cbind(y, y * yhat, yhat^2)
    Sig <- cov(Z)
    c1 <- matrix(c((V2 - V1 + 2 * pp * V1) / V2^2, -2 / V2, 1 / V2), nrow = 1)
    estsd <- sqrt((c1 %*% Sig %*% t(c1))[1])
  }
  zstat <- (rx2 - ry2) / estsd
  pval_xy <- 1 - pnorm(zstat)

  # CIs <- c(rx2 - ry2 + qnorm(alpha/2) * estsd, rx2 - ry2 + qnorm(1- alpha/2) * estsd)
  return(list(test.stat = zstat, pval = pval_xy))
}

Steiger <- function(df1, df2) {
  n1 <- nrow(df1)
  n2 <- nrow(df2)
  p <- ncol(df1) 
  
  # select the SNP for test statistic construction
  rx <- cor(df1$X, df1[, 3:p])
  ry <- cor(df2$Y, df2[, 3:p])
  rxy <- abs(rx) + abs(ry)
  idx <- which.max(rxy)
  
  # construct the test statistics
  zx <- 0.5 * log((1 + abs(rx[idx]))/(1 - abs(rx[idx])))
  zy <- 0.5 * log((1 + abs(ry[idx]))/(1 - abs(ry[idx])))
  z <- (zx - zy) / sqrt(1 / (n1 - 3) + 1 / (n2 - 3))
  pval_xy <- 1 - pnorm(z)
  
  return(list(test.stat = z, pval = pval_xy))
}

```


```{r}
## one testing 
simu1_func <- function(oriSNP, uhat.sd, num.sample, sample.type = 'sample', 
                       stage1_mdl, stage2_mdl, stage2_mdl_null, sigma_Y, 
                       effect.size.U, effect.size.X, split.r, setseed = 1025,
                       mdl1.type = "LM",  ori.mdl2 = "LM", mdl2.type = "LM-RI", test.type, num.pseudo = 10000) {
  # data generation
  df_lst <- genData(oriSNP, uhat.sd, num.sample, sample.type, mdl2.type, sigma_Y, 
                    stage1_mdl, ori.mdl2, stage2_mdl, stage2_mdl_null, effect.size.U, effect.size.X, split.r, setseed)
  XtoY.df.1 <- df_lst[[1]][[1]]
  XtoY.df.2 <- df_lst[[1]][[2]]
  YtoX.df.1 <- df_lst[[2]][[1]]
  YtoX.df.2 <- df_lst[[2]][[2]]
  XtoY.n1 <- nrow(XtoY.df.1)
  XtoY.n2 <- nrow(XtoY.df.2)
  YtoX.n1 <- nrow(YtoX.df.1)
  YtoX.n2 <- nrow(YtoX.df.2)
  
  for (i in 1:2) {
    if (i == 1) {
      df1 <- XtoY.df.1
      df2 <- XtoY.df.2
      n1 <- XtoY.n1
      n2 <- XtoY.n2
    }else{
      df1 <- YtoX.df.1
      df2 <- YtoX.df.2
      n1 <- YtoX.n1
      n2 <- YtoX.n2
    }
    
    # model fitting
    ## 1st stage
    if (mdl1.type == "LM") {
      lm1.formula <- paste0("X ~ ", paste0(colnames(df1)[grepl("SNP", colnames(df1))], collapse = " + "))
      lm1.formula <- as.formula(lm1.formula)
      mdl1 <- lm(lm1.formula, data = df1)
    }else{
      print("Method not implemented. ")
    }
    
    SNPs <- df1[, 3:(ncol(df1))]
    M <- ncol(df1) - 2
    # Rx2 <- matrix(cor(df1$X, SNPs), nrow = 1) %*% solve(cor(SNPs, SNPs), tol = 1e-40) %*% matrix(cor(df1$X, SNPs), ncol = 1)
    Rx2 <- summary(mdl1)$r.squared
    lbdx <- n1 * Rx2 / (1 - Rx2)
    
    ## 2nd stage
    mdl2.type <- strsplit(mdl2.type, "-")[[1]]
    SNPs2 <- df2[, 3:(ncol(df2))]
    if (mdl2.type[2] == "RI") {
      uhat <- df2$Y - predict(mdl1, newdata = df2)
      if (mdl2.type[1] == "LM") {
        mdl2 <- lm(df2$Y ~ df2$X + uhat)
        Ry2.1 <- as.numeric(matrix(cor(df2$Y, SNPs2), nrow = 1) %*% solve(cor(SNPs2, SNPs2), tol = 1e-30) %*% matrix(cor(df2$Y, SNPs2), ncol = 1))
        Ry2.2 <- cor(df2$Y, df2$X)^2 / cor(df2$X, df2$X)
        # Ry2 <- summary(mdl2)$r.squared
        yhat <- mdl2$fitted.values - mdl2$coefficients[3] * uhat
        newY <- df2$Y - mdl2$coefficients[3] * uhat
      }else{
        mdl2 <- glm(df2$Y ~ df2$X + uhat, family = binomial(link = "logit"))
        # mdl2.2 <- glm(df2$Y ~ uhat, family = binomial(link = "logit"))
        # mdl2 <- glm(df2$Y ~ df2$X, family = binomial(link = "logit"))
        # yhat <- mdl2$fitted.values
        yhat <- 1 / (1 + exp( - mdl2$coefficients[1] - mdl2$coefficients[2] * df2$X))
        Ry2 <- glmR2_func(df2$Y, yhat)[2]
        # newY <- ?
        # Ry2 <- summary(mdl2.1)$r.squared - summary(mdl2.2)$r.squared
      }
    }else if (mdl2.type[2] == "PS") {
      xhat <- predict(mdl1, newdata = df2)
      if (mdl2.type[1] == "LM") {
        mdl2 <- lm(df2$Y ~ xhat)
        # Ry2 <- as.numeric(matrix(cor(df2$Y, SNPs2), nrow = 1) %*% solve(cor(SNPs2, SNPs2), tol = 1e-30) %*% matrix(cor(df2$Y, SNPs2), ncol = 1))
        Ry2 <- summary(mdl2)$r.squared
        yhat <- mdl2$fitted.values
        newY <- df2$Y
      }else{
        mdl2 <- glm(df2$Y ~ xhat, family = binomial(link = "logit"))
        yhat <- mdl2$fitted.values
        Ry2 <- glmR2_func(df2$Y, yhat)[2]
        newY <- df2$Y
      }
    }
    lbdy <- n2 * Ry2 / (1 - Ry2)
    
    # hypothesis testing
    if(length(test.type) == 1) {
      test_res <- switch (test.type,
                          "F" = fTest(Rx2, Ry2, M, n1, n2, lbdx, lbdy, num.pseudo), 
                          "Beta" = betaTest(Rx2, Ry2, M, n1, n2, lbdx, lbdy, num.pseudo), 
                          "R2" = R2Test(Rx2, Ry2, M, n1, n2, alpha, mdl2.type[1], newY, yhat),
                          "Steiger" = Steiger(df1, df2)
      )
      test.res <- c(test_res$test.stat, test_res$pval)
      res.cols <- c(paste0(test.type, "test.stat"), paste0(test.type, "pval_xy"))
    }else{
      test_res <- rep(NA, 2 * length(test.type))
      res.cols <- c(sapply(test.type, function(x) paste(x, c("test.stat", "pval_xy"))))
      for (k in 1:length(test.type)) {
        test_resk <- switch (test.type[k],
                             "F" = fTest(Rx2, Ry2, M, n1, n2, lbdx, lbdy, num.pseudo), 
                             "Beta" = betaTest(Rx2, Ry2, M, n1, n2, lbdx, lbdy, num.pseudo),
                             "R2" = R2Test(Rx2, Ry2, M, n1, n2, alpha, mdl2.type[1], df2$Y, yhat), 
                             "Steiger" = Steiger(df1, df2)
                             )
        test_res[(k - 1) * 2 + 1] = test_resk$test.stat
        test_res[(k - 1) * 2 + 2] = test_resk$pval
      }
    }
  }
  return(list(r.squared = c(Rx2, Ry2), lambda = c(lbdx, lbdy), testRes = test_res, res.cols = res.cols))
}



## the complete simulation function

simu_func <- function(M, oriSNP, uhat.sd, num.sample, sample.type = 'sample', 
                      stage1_mdl, stage2_mdl, stage2_mdl_null, sigma_Y, 
                      effect.size.U, effect.size.X, split.r, setseed = 1025, 
                      mdl1.type = "lm", mdl2.type = "LM-RI", test.type = "R2", num.pseudo = 10000) {
  
  res <- pblapply(1:M, function(i) simu1_func(oriSNP, uhat.sd, num.sample, sample.type, 
                                            stage1_mdl, stage2_mdl, stage2_mdl_null, sigma_Y, 
                                            effect.size.U, effect.size.X, split.r, 
                                            setseed = setseed * i + 1992,
                                            mdl1.type,  mdl2.type, test.type, num.pseudo))
  r.squared <- t(sapply(res, function(x) x$r.squared)) # M x 2
  lambda.xy <- t(sapply(res, function(x) x$lambda)) # M x 2
  test.res <- t(sapply(res, function(x) x$testRes))
  colnames(test.res) <- res[[1]]$res.cols
  return(list(r.squared = r.squared, 
              lambda.xy = lambda.xy, 
              test.res = test.res))
}
```

```{r simulation I data generation I}

#####################  Load packages  #####################
source("D:/Projects/Casual-Rsquared/Simulation code/code/utils.R")
library(hypergeo)
library(pbapply)
pboptions(type = "txt", char = "=")

#####################  Load sample dataset  #####################

load("D:/Projects/Casual-Rsquared/twas_methods-master/sample_gene_expression_and_snp.Rdata")

colnames(cleaning_result)[4:33] = paste("SNP",1:30,sep="")
colnames(cleaning_result)[2] <- "y"
colnames(cleaning_result)[3] <- "X"

lm_formula = paste("X ~ ", paste("SNP", 1:30, sep = "",collapse =" + "))
lm_formula = as.formula(lm_formula)
lm1 = step(lm(lm_formula,data = cleaning_result), direction = "backward", trace = -1)
# include.idx <- sapply(strsplit(rownames(summary(lm1)$coefficients), "SNP")[-1], function(x) as.numeric(x[2]))

# Predict y and u
X_hat = predict(lm1)
u_hat = cleaning_result$X - X_hat
uhat.sd = sd(u_hat)

# # Linear regression for stage 2
lm2.1 <- lm(cleaning_result$y ~ u_hat)
lm2.2 <- lm(cleaning_result$y ~ cleaning_result$X + u_hat)

# Logistic Regression Only u_hat
glm2.1 = glm(cleaning_result$y ~ u_hat, family = binomial(link = "logit"))

# Logistic Regression with y and u_hat
glm2.2 = glm(cleaning_result$y ~ cleaning_result$X + u_hat,
             family = binomial(link = "logit"))

#####################  Parameter Settings  #####################


num.replicates <- c(1, 3, 5, 10)
effect.size.X <- c(0, 1, 3, 5, 10)
effect.size.U <- c(0, 1, 3, 5, 10, 30, 50)
split.r <- 1 # c(1, 2, 3, 5, 10)
sigma.Y <- 1 # c(0, 0.05, 0.1, 0.2, 0.5, 1, 2)
M <- 1000 # replicates
stage1_mdl <- lm1 
mdl1.type <- "LM"
# mdl2.type <- c("LM-RI", "LM-PS", "GLM-RI", "GLM-PS")
mdl2.type <- c("LM-RI", "LM-PS", "GLM-PS")
stage2_mdl <- list(lm2.2, lm2.2, glm2.2)
stage2_mdl_null <- list(lm2.1, lm2.1, glm2.1)
# stage2_mdl <- glm2.2
# stage2_mdl_null <- glm2.1
oriSNP <- cleaning_result[, 4:33]
sample.type = 'sample'
num.pseudo <- 10000
setseed <- 1025
test.type <- list(c("F", "R2", "Steiger"), 
                  c("F", "R2", "Steiger"), 
                  c("R2"))

#####################  Simulation  #####################

# sample size
res <- vector("list", length(num.replicates))
for (i in 1:length(num.replicates)) {
  # effect.size.X
  res[[i]] <- vector("list", length(effect.size.X))
  for (k in 1:length(effect.size.X)) {
    # effect.size.U
    res[[i]][[k]] <- vector("list", length(effect.size.U))
    for (m in 1:length(effect.size.U)) {
      # mdl2.type
      res[[i]][[k]][[m]] <-  vector("list", length(mdl2.type))
      for (p in 1:length(mdl2.type)) {
          res[[i]][[k]][[m]][[p]] <- simu_func(M, oriSNP, uhat.sd, num.replicates[i] * nrow(oriSNP), sample.type, 
                                              stage1_mdl, stage2_mdl[[p]], stage2_mdl_null[[p]], sigma.Y, 
                                              effect.size.U[m], effect.size.X[k], split.r, setseed, 
                                              mdl1.type, mdl2.type[p], test.type[[p]], num.pseudo)
          }
        }
      }
    }
  }
}
