---
title: "Simulated real data simulation"
output: html_document
date: "2022-11-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(hypergeo)
```

## True model
$$X_i = \beta_{X0} + \sum_{m = 1}^M \beta_{Xm}SNP_{i,m} + \mathbf{U}_i\boldsymbol{\beta}_{XU} + \epsilon_i = \beta_{X0} +  \mathbf{G}_i\boldsymbol{\beta}_{Xg} +  \mathbf{U}_i\boldsymbol{\beta}_{XU} + \epsilon_i$$
GLM for the second stage as follows: 
$$logit(p_i) = \alpha_{y0} + \alpha_{yx}X_i + \boldsymbol{\alpha}_{yU}\mathbf{U}_i$$
where $p_i = \text{Prob}(y_i = 1)$ or LM
$$y_i = \alpha_{y0} + \alpha_{yx}X_i + \boldsymbol{\alpha}_{yU}\mathbf{U}_i$$
with SNPs taken as IVs satisfying assumptions as in MR: 
* IVs are associated with response $X$
* IVs are not associated with confounders $U$
* Conditional on $X$, response $y$ and IVs are not associated

Let's start with the ideal situation when all the IVs are valid IVs, and all these assumptions are met. 

```{r Different tests}
## beta test
beta_mu_var <- function(alp, bt, lbd){
  uu <- c(alp + bt, alp + 1)
  ll <- c(alp, alp + bt + 1)
  z <- lbd/2
  # mu <- exp(-lbd/2) * gamma(alp + 1) / gamma(alp) * gamma(alp + bt) / gamma(alp + bt + 1) * genhypergeo(uu, ll, z)
  mu <- exp(-lbd/2) * alp / (alp + bt) * genhypergeo(uu, ll, z)
  uu <- c(alp + bt, alp + 2)
  ll <- c(alp, alp + bt + 2)
  sigma <- exp(-lbd/2) * alp * (alp + 1) / ((alp + bt) * (alp + bt + 1)) * genhypergeo(uu, ll, z) - mu^2
  if(exp(-lbd/2) == 0) {
    mu <- 0
    sigma <- 0
  }
  return(c(mu, sigma))
}

betaTest <- function(rxg2, ryg2, nSNP, n1, n2, lambda_x, lambda_y, alpha) {
  var1 <- beta_mu_var(nSNP/2, (n1 - nSNP - 1)/2, lambda_x)[2]
  var2 <- beta_mu_var(nSNP/2, (n2 - nSNP - 1)/2, lambda_y)[2]
  zstat <- (rxg2 - ryg2) / sqrt(var1 + var2)
  pval0 <- pnorm(zstat)
  CIs <- c(rxg2 - ryg2 + qnorm(alpha/2) * sqrt(var1 + var2), rxg2 - ryg2 + qnorm(1- alpha/2) * sqrt(var1 + var2))
  return(list(statistic = zstat, pval = pval0, CI = CIs))
}

## f test
f_mu_var <- function(n, m, lambda){
  if(m > 2){
    mu <- m * (n + lambda) / (n * (m - 2))
    if(m > 4){
      sigma <- 2 * ((n + lambda)^2 + (n + 2 * lambda) * (m - 2))/((m - 2)^2 * (m - 4)) * (m / n)^2
    }else{
      sigma <- NA
    }
  }else{
    mu <- NA
    sigma <- NA
  }
  return(c(mu, sigma))
}


fTest <- function(rxg2, ryg2, nSNP, n1, n2, lambda_x, lambda_y, alpha, M = 10000) {
  ## construct empirical distribution
  nu <- 0.5 * ((nSNP + lambda_x) / (n1 - nSNP - 3) + (nSNP + lambda_y) / (n2 - nSNP - 3))
  
  Tx <- rf(M, nSNP, n1, max(0, nu * (n1 - nSNP - 3) - nSNP))
  Ty <- rf(M, nSNP, n2, max(0, nu * (n2 - nSNP - 3) - nSNP))
  
  pval <- mean(rxg2 - ryg2 > Tx - Ty)

  return(list(statistic = rxg2 - ryg2, pval = pval))
}

## when n sufficiently large
chi2Test <- function(rxg2, ryg2, nSNP, lambda_x, alpha) {
  fstat <- rxg2/ryg2
  pval0 <- pf(fstat, nSNP, nSNP, lambda_x)
  upper_crit <- 1/qf(alpha/2, nSNP, nSNP, lambda_x)
  lower_crit <- qf(alpha/2, nSNP, nSNP, lambda_x)
  CIs <- c(fstat * lower_crit, fstat * upper_crit)
  return(list(statistic = fstat, pval = pval0, CI = CIs))
}


## R2 test
R2Test <- function(rxg2, ryg2, nSNP, n1, n2, alpha, type, y, phat) {
  if(type == "lm") {
    var1 <- 4 * rxg2 * (1 - rxg2)^2 * (n1 - nSNP - 1)^2 / ((n1^2 - 1) * (n1 + 3))
    var2 <- 4 * ryg2 * (1 - ryg2)^2 * (n2 - nSNP - 1)^2 / ((n2^2 - 1) * (n2 + 3))
    estsd <- sqrt(var1 + var2)
  }else{
    pp <- mean(phat)
    V1 <- mean(phat * (1 - phat))
    V2 <- var(y)
    Z <- cbind(y, y * phat, phat^2)
    Sig <- cov(Z)
    c1 <- matrix(c(V2 - V1 + 2 * pp * V1, -2 / V2, 1 / V2), nrow = 1)
    estsd <- sqrt((c1 %*% Sig %*% t(c1))[1])
  }
  zstat <- (rxg2 - ryg2) / estsd
  pval0 <- pnorm(zstat)
  CIs <- c(rxg2 - ryg2 + qnorm(alpha/2) * estsd, rxg2 - ryg2 + qnorm(1- alpha/2) * estsd)
  return(list(statistic = zstat, pval = pval0, CI = CIs))
}

```

```{r}
glmR2_func <- function(y, phat) {
  SST <- sum((y - mean(y))^2)
  SSE <- sum((y - phat)^2)
  Rss2 <- 1 - SSE / SST
  Rr2 <- sum((y - mean(y)) * (phat - mean(y)))^2 / (sum((y - mean(y))^2) * sum((phat - mean(y))^2))
  return(c(Rss2, Rr2))
}

```

## Two-sample 

### Haoran's small dataset
```{r}
load("twas_methods-master/sample_gene_expression_and_snp.Rdata")

colnames(cleaning_result)[4:33] = paste("SNP",1:30,sep="")
colnames(cleaning_result)[2] <- "y"
colnames(cleaning_result)[3] <- "X"

lm_formula = paste("X ~ ", paste("SNP", 1:30, sep = "",collapse =" + "))
lm_formula = as.formula(lm_formula)
lm1 = step(lm(lm_formula,data = cleaning_result), direction = "backward", trace = -1)
include.idx <- sapply(strsplit(rownames(summary(lm1)$coefficients), "SNP")[-1], function(x) as.numeric(x[2]))

# Predict y and u
X_hat = predict(lm1)
u_hat = cleaning_result$X - X_hat
uhat.sd = sd(u_hat)

# # Linear regression for stage 2
# lm2.1 <- lm(cleaning_result$y ~ u_hat)
# lm2.2 <- lm(cleaning_result$y ~ cleaning_result$X + u_hat)
  
# Logistic Regression Only u_hat
glm2.1 = glm(cleaning_result$y ~ u_hat, family = binomial(link = "logit"))

# Logistic Regression with y and u_hat
glm2.2 = glm(cleaning_result$y ~ cleaning_result$X + u_hat,
           family = binomial(link = "logit"))
```

```{r split function}
split_R2_test <- function(M, num.replicates, effect.size.X, effect.size.U, split.r, 
                          stage1_mdl, stage2_mdl, stage2_mdl_null, include.idx, 
                          uhat.sd, dataset, SNP_idx, seed = 1025, alpha = 0.05) {
  # set seed for replicate
  set.seed(seed)
  n0 <- nrow(dataset)
  nSNP <- length(SNP_idx)
  SNP0 <- dataset[, SNP_idx]
  num.sample <- n0 * num.replicates
  n1 <- floor(num.sample / (1 + split.r))
  n2 <- num.sample - n1
  idx1 <- sample(num.sample, n1)
  idx2 <- setdiff(1:num.sample, idx1)
  
  beta_all <- NULL
  f_all <- NULL
  chisq_all <- NULL
  R2_lm_all <- NULL
  R2_glm_all <- NULL
  
  for (i in 1:M) {
    # sample U ~ N(0, hat_u_sd)
    U <- rnorm(num.sample, sd = uhat.sd)
    
    # sample with replacement to get SNPs
    SNPs_idx <- sample(n0, num.sample, replace = TRUE)
    SNPs <- SNP0[SNPs_idx, ]
    
    # generate X
    X <- as.matrix(cbind(1, SNPs[, include.idx])) %*% matrix(stage1_mdl$coefficients, ncol = 1) + U
    
    # generate y
    if (effect.size.X == 0) {
      logit.p <- cbind(1, U) %*% matrix(stage2_mdl_null$coefficients * c(1, effect.size.U), ncol = 1)
    }else{
      logit.p <- cbind(1, X, U) %*% matrix(stage2_mdl$coefficients * c(1, effect.size.X, effect.size.U), ncol = 1)
    }
    p <- exp(logit.p) / (1 + exp(logit.p))
    y <- rbinom(num.sample, 1, p)
    
    # get datasets for two stages
    simulated_data <- data.frame(y = y, X = X, SNP = SNPs)
    colnames(simulated_data)[3:(2 + nSNP)] <- colnames(SNP0)
    stage1.df <- simulated_data[idx1, ]
    stage2.df <- simulated_data[idx2, ]
    
    # First stage
    lm1 <- lm(stage1_mdl$call$formula, data = stage1.df)
    Xhat <- predict(lm1, newdata = stage2.df)
    uhat <- stage2.df$X - Xhat
    Rxg2.1 <- summary(lm1)$r.squared
    Rxg2.2 <- (cor(stage1.df$X, stage1.df[, 3:(2 + nSNP)]) %*% solve(cor(stage1.df[, 3:(2 + nSNP)]), tol = 1e-35) %*% t(cor(stage1.df$X, stage1.df[, 3:(2 + nSNP)])))[1]
    lambda_x <- sum(Xhat^2)/var(stage1.df$X)

    # second stage
    ## linear model
    lm_2sps <- lm(stage2.df$y ~ Xhat)
    lm_2sri <- lm(stage2.df$y ~ stage2.df$X + uhat)
    lm_2sri_2 <- lm(stage2.df$y ~ Xhat + uhat)
    Ryg2.lm.1 <- rep(NA, 3)
    SST <- sum((stage2.df$y - mean(stage2.df$y))^2)
    yhat1 <- lm_2sps$fitted.values
    yhat2 <- lm_2sri$fitted.values - uhat * lm_2sri$coefficients[3]
    yhat3 <- lm_2sri_2$fitted.values - uhat * lm_2sri_2$coefficients[3]
    SSE1 <- sum((stage2.df$y - mean(yhat1))^2)
    SSE2 <- sum((stage2.df$y - mean(yhat2))^2)
    SSE3 <- sum((stage2.df$y - mean(yhat3))^2)
    Ryg2.lm.1 <- 1 - c(SSE1/SST, SSE2/SST, SSE3/SST)
    Ryg2.lm.2 <- tryCatch((cor(stage2.df$y, stage2.df[, 3:(2 + nSNP)]) %*% 
      solve(cor(stage2.df[, 3:(2 + nSNP)]), tol = 1e-35) %*% t(cor(stage2.df$y, stage2.df[, 3:(2 + nSNP)])))[1], 
      error=function(e){cat("ERROR :", conditionMessage(e), "\n")})
    denum_term <- var(stage2.df$y) + var(stage1.df$X) * c(lm_2sps$coefficients[2]^2, lm_2sri$coefficients[2]^2, lm_2sri_2$coefficients[2]^2)
    lambda_y <- sum(Xhat^2) * c(lm_2sps$coefficients[2]^2, lm_2sri$coefficients[2]^2, lm_2sri_2$coefficients[2]^2)
    lambda_y <- lambda_y / denum_term
    
    ## logistic model
    glm_2sps <- glm(stage2.df$y ~ Xhat, family = binomial(link = "logit"))
    glm_2sri <- glm(stage2.df$y ~ stage2.df$X + uhat, family = binomial(link = "logit"))
    glm_2sri_2 <- glm(stage2.df$y ~ Xhat + uhat, family = binomial(link = "logit"))
    phat <- matrix(NA, nrow = length(idx2), ncol = 3)
    phat.1 <- cbind(1, Xhat) %*% matrix(glm_2sps$coefficients, ncol = 1)
    phat.2 <- 1 + phat.1
    phat[, 1] <- phat.1 / phat.2
    phat.1 <- cbind(1, stage2.df$X) %*% matrix(glm_2sri$coefficients[1:2], ncol = 1)
    phat.2 <- 1 + phat.1
    phat[, 2] <- phat.1 / phat.2
    phat.1 <- cbind(1, Xhat) %*% matrix(glm_2sri_2$coefficients[1:2], ncol = 1)
    phat.2 <- 1 + phat.1
    phat[, 3] <- phat.1 / phat.2
    Ryg2.glm <- sapply(1:3, function(i) glmR2_func(stage2.df$y, phat[, i])) # 2 x 3 matrix
    
    ## Beta Testing results
    
    mdl_nms <- rep(c("2sps", "2sri", "2sri2"), each = 2)
    R2_names <- rep(c("R1.R1.", "R2.R1"), 3)
    beta_names <- paste(mdl_nms, R2_names, sep = ".")
    beta_res <- NULL
    for (i in 1:3) {
      beta_res_1 <- betaTest(Rxg2.1, Ryg2.lm.1[i], nSNP, n1, n2, lambda_x, lambda_y[i], alpha)$pval
      beta_res_2 <- betaTest(Rxg2.2, Ryg2.lm.1[i], nSNP, n1, n2, lambda_x, lambda_y[i], alpha)$pval
      # beta_res_3 <- betaTest(Rxg2.2, Ryg2.lm.2, nSNP, n1, n2, lambda_x, lambda_y[i], alpha)$pval
      beta_res <- c(beta_res, beta_res_1, beta_res_2) # , beta_res_3)
    }
    
    beta_all <- rbind(beta_all, beta_res)
    
    ## F testing results
    mdl_nms <- rep(c("2sps", "2sri", "2sri2"), each = 2)
    R2_names <- rep(c("R1.R1.", "R2.R1"), 3)
    f_names <- paste(mdl_nms, R2_names, sep = ".")
    f_res <- NULL
    for (i in 1:3) {
      f_res_1 <- fTest(Rxg2.1, Ryg2.lm.1[i], nSNP, n1, n2, lambda_x, lambda_y[i], alpha)$pval
      f_res_2 <- fTest(Rxg2.2, Ryg2.lm.1[i], nSNP, n1, n2, lambda_x, lambda_y[i], alpha)$pval
      # f_res_3 <- fTest(Rxg2.2, Ryg2.lm.2, nSNP, n1, n2, lambda_x, lambda_y[i], alpha)$pval
      f_res <- c(f_res, f_res_1, f_res_2) #, f_res_3)
    }
    f_all <- rbind(f_all, f_res)
    
    
    ## chi-squared results
    mdl_nms <- rep(c("2sps", "2sri", "2sri2"), each = 2)
    R2_names <- rep(c("R1.R1.", "R2.R1"), 3)
    chisq_names <- paste(mdl_nms, R2_names, sep = ".")
    # chisq_names <- c(chisq_names, "R2.R2")
    chisq_res <- NULL
    for (i in 1:3) {
      chisq_res_1 <- chi2Test(Rxg2.1, Ryg2.lm.1[i], nSNP, lambda_x, alpha)$pval
      chisq_res_2 <- chi2Test(Rxg2.2, Ryg2.lm.1[i], nSNP, lambda_x, alpha)$pval
      chisq_res <- c(chisq_res, chisq_res_1, chisq_res_2)
    }
    chisq_all <- rbind(chisq_all, chisq_res)
    # chisq_res <- c(chisq_res, chi2Test(Rxg2.2, Ryg2.lm.2, nSNP, lambda_x, alpha)$pval)
    
    ## R2 test results
    R2_res_lm.1 <- sapply(1:3, function(i) R2Test(Rxg2.1, Ryg2.lm.1[i], nSNP, n1, n2, alpha, "lm", stage2.df$y, phat[, i])$pval)
    R2_res_lm.2 <- sapply(1:3, function(i) R2Test(Rxg2.2, Ryg2.lm.1[i], nSNP, n1, n2, alpha, "lm", stage2.df$y, phat[, i])$pval)
    # R2_res_lm.3 <- sapply(1:3, function(i) R2Test(Rxg2.2, Ryg2.lm.2, nSNP, n1, n2, alpha, "lm", stage2.df$y, phat[, i])$pval)
    R2_res_glm <- sapply(1:3, function(i) sapply(1:2, function(j) R2Test(Rxg2.1, Ryg2.glm[j, i], nSNP, n1, n2, alpha, "glm", stage2.df$y, phat[, i])$pval))
    R2_lm_all <- rbind(R2_lm_all, c(R2_res_lm.1, R2_res_lm.2))
    R2_glm_all <- rbind(R2_glm_all, as.numeric(R2_res_glm))
  }
  colnames(beta_all) <- beta_names
  colnames(f_all) <- f_names
  colnames(chisq_all) <- chisq_names
  colnames(R2_lm_all) <- paste(rep(c("2sps", "2sri", "2sri2"), 2), rep(c("R1.R1.", "R2.R1"), each = 3), sep = ".")
  finalResults <- list(beta = beta_all, 
                        f = f_all, 
                        chisq = chisq_all, 
                        R2.lm = R2_lm_all,
                        R2.glm = R2_glm_all)
  return(finalResults)
}

```

```{r}
num.replicates <- c(1, 3, 5, 10, 15, 20)
effect.size.X<- c(0, 1, 2, 3, 5, 10)
effect.size.U <- c(1, 10, 20, 30, 40, 50)
split.r <- c(1, 2, 3, 5, 10)
M <- 100
stage1_mdl <- lm1
stage2_mdl <- glm2.2
stage2_mdl_null <- glm2.1
dataset <- cleaning_result
SNP_idx <- 4:33

res <- vector("list", length(num.replicates))
for (i in 1:length(num.replicates)) {
  cat("Num.replicates = ", num.replicates[i], "\n")
  res[[i]] <- vector("list", length(split.r))
  for (j in 1:length(split.r)) {
    cat("split ratio = ", split.r[j], "\n")
    res[[i]][[j]] <- vector("list", length(effect.size.X))
    for (pp in 1:length(effect.size.X)) {
      cat("effect.size.X = ", effect.size.X[pp], "\n")
      res[[i]][[j]][[pp]] <- vector("list", length(effect.size.U))
      for (qq in 1:length(effect.size.U)) {
        cat("effect.size.U = ", effect.size.X[qq], "\n")
        res[[i]][[j]][[pp]][[qq]] <- split_R2_test(M, num.replicates[i], effect.size.X[pp], effect.size.U[qq], split.r[j], 
                                                   stage1_mdl, stage2_mdl, stage2_mdl_null, include.idx, 
                                                   uhat.sd, dataset, SNP_idx, seed = 1025, alpha = 0.05)
      }
    }
  }
  
}

```

```{r}
par(mfrow = c(1, 4))
for (i in 1:length(num.replicates)) {
  for (j in 1:length(split.r)) {
    for (pp in 1:length(effect.size.X)) {
      for (qq in 1:length(effect.size.U)) {
        if(!is.null(resi)) {
          cat("Num.replicates = ", num.replicates[i], "\n")
          cat("split ratio = ", split.r[j], "\n")
          cat("effect.size.X = ", effect.size.X[pp], "\n")
          cat("effect.size.U = ", effect.size.X[qq], "\n")
          R2_glm <- paste(rep(c("2sps", "2sri", "2sri2"), 2), rep(c("Rss2.", "Rr2"), each = 3), sep = ".")
          for (mm in 1:6) {
            # print(hist(resi$beta[, mm], xlab = "pvals under beta", main = colnames(resi$beta)[mm]))
            print(hist(resi$f[, mm], xlab = "pvals under f", main = colnames(resi$f)[mm]))
            print(hist(resi$chisq[, mm], xlab = "pval under chisq", main = colnames(resi$chisq)[mm]))
            print(hist(resi$R2.lm[, mm], xlab = "pval under R2.lm", main = colnames(resi$R2.lm)[mm]))
            print(hist(resi$R2.glm[, mm], xlab = "pval under R2.lm", main = colnames(resi$R2.lm)[mm]))
          }
        }
      }
    }
  }
}

```


```{r}
resi <- res[[1]][[1]][[1]][[1]]
par(mfrow = c(1, 4))
for (mm in 1:6) {
            # print(hist(resi$beta[, mm], xlab = "pvals under beta", main = colnames(resi$beta)[mm]))
            print(hist(resi$f[, mm], xlab = "pvals under f", main = colnames(resi$f)[mm]))
            print(hist(resi$chisq[, mm], xlab = "pval under chisq", main = colnames(resi$chisq)[mm]))
            print(hist(resi$R2.lm[, mm], xlab = "pval under R2.lm", main = colnames(resi$R2.lm)[mm]))
            print(hist(resi$R2.glm[, mm], xlab = "pval under R2.glm", main = colnames(resi$R2.lm)[mm]))}
```

```{r}
resi <- res[[1]][[1]][[2]][[1]]
lapply(resi, function(x) summary(x))
par(mfrow = c(1, 4))
for (mm in 1:6) {
            #print(hist(resi$beta[, mm], xlab = "pvals under beta", main = colnames(resi$beta)[mm]))
            print(hist(resi$f[, mm], xlab = "pvals under f", main = colnames(resi$f)[mm]))
            print(hist(resi$chisq[, mm], xlab = "pval under chisq", main = colnames(resi$chisq)[mm]))
            print(hist(resi$R2.lm[, mm], xlab = "pval under R2.lm", main = colnames(resi$R2.lm)[mm]))
            print(hist(resi$R2.glm[, mm], xlab = "pval under R2.glm", main = colnames(resi$R2.lm)[mm]))}
```

```{r}
resi <- res[[1]][[1]][[1]][[1]]
lapply(resi, function(x) summary(x))

resi <- res[[1]][[1]][[2]][[1]]
lapply(resi, function(x) summary(x))

resi <- res[[1]][[1]][[3]][[1]]
lapply(resi, function(x) summary(x))

resi <- res[[1]][[1]][[4]][[1]]
lapply(resi, function(x) summary(x))

resi <- res[[1]][[1]][[5]][[1]]
lapply(resi, function(x) summary(x))

resi <- res[[1]][[1]][[6]][[1]]
lapply(resi, function(x) summary(x))
```

### Ruoyu's dataset
```{r}



```